Network based multifactorial modelling microRNA-target interaction.

### 2019-08-09 article.tex
overview

  \title{Network based multifactorial modelling microRNA-target interaction.}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Selcen Ari\footnote{Yildiz Technical University Department of
  Bioengineering} \\ Alper Yilmaz\footnote{Yildiz Technical University Department of
  Bioengineering}}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{28 Mayıs 2019}

@@

Competing endogenous RNA (ceRNA) regulations and crosstalk between
various types of non-coding RNA in human is remarkable in means of miRNA
regulation. Many studies have pointed out that an alteration in
miRNA:target interaction can result in unexpected changes due to
indirect and complex interactions. In this paper, we defined a new
network-based model that handles miRNA:ceRNA interactions with
expression values. Our model is able to handle miRNA interaction factors
such as seed type, binding energy, if provided. Our approach is able to
reveal that a perturbation in an element of network affects whole
competing elements differently and cooperative efficiencies of miRNAs on
common targets could be calculated. Our findings emphasized importance
of miRNA:target ratios being crucial, as reported by previous studies.
We have showed that the competing elements which have the same or close
expression values may not be affected equally from the perturbation
because of repression functionality depended on interaction factors of
miRNA target pairs (TOFIX: Hocam etkileşim faktörlerinden dolayı
ekspresyonları yakın olsa bile baskılanma aktitesinin aynı
sergilenmeyeceğini ifade etmeye çalıştım. Bu şekilde anlaşılıyor mu?).
We applied the model to real sample consisting of breast cancer gene and
miRNA expression dataset and experimental miRNA:target interaction
dataset all generated via high throughput sequencing methods. A gene
over-expressed in tumor tissue, namely \emph{ABCC1}, is used as
perturbing element. We have observed that change in expression level of
single gene in miRNA:target network is sufficient to perturb regulations
in whole network, due to unforeseen and unpredicted regulation which are
only visible when considered in network context. Therefore, this model
helps unveiling the crosstalk between elements in miRNA:target network
where abundance of target and sponge effect are taken into account. The
model is scalable and can be plugged in with emerging miRNA effectors
such as circRNAs. The model is available as R package at
\url{https://github.com/selcenari/ceRNAnetsim}. (TODO the link will
change)

@@

As it is possible for miRNAs to suppress multiple targets, an individual
mRNA molecule can also be targeted by multiple miRNAs. In that case, the
targeted mRNAs exhibit competitor behavior, that is hypothesized as
competing endogenous RNAs (ceRNAs) (Ala et al. 2013; Cesana and Daley
2013), against their miRNAs. Briefly, Ala et al.~have explained the
ceRNA hypothesis as disturbance of the other target when one of the
targets on a steady-state system that included one miRNA and two target
was perturbed with expression change (Ala et al. 2013). Regarding
interaction between miRNAs and their target in a cell, explaining and
predicting results of an individual perturbation is difficult due to
complexity of interactions. Various computational and experimental
studies have tackled the problem of unraveling ceRNA:miRNA interactions.
It has been observed that when abundance of one of the targets of
miR-122 was increased, the other target expression also slightly
increased as a result of decreasing repressive activity of miR-122 on
them (Denzler et al. 2014). Bosson et al.~have developed a mathematical
model for changes on total target pool concentration after grouping
targets according to affinity and demonstrated that miRNA activity
correlated with affinity between miRNA and target (Bosson, Zamudio, and
Sharp 2014). Cooperative efficiency of miRNAs as well as competitor
behaviors of targets were also studied and it has been demonstrated to
be crucial for regulating available mRNA levels of targets (Denzler et
al. 2016). MiRNA:target interactions have been modeled as stoichiometric
and catalytic mechanisms and Figliuzzi et al.~have recommended handling
models in network context (Figliuzzi, Marinari, and De Martino 2013).
The model that can explain miRNA target interaction through topological
features has been applied at bipartite network by Nitzan et al.~(Nitzan
et al. 2014). Robinson and Henderson applied the model that handles
miRNA:target direct and indirect interactions via common miRNA of gene
and target of miRNA, at bipartite network. It has been demonstrated that
all miRNAs and targets in the network can interact with each other
through common miRNAs and genes, without interaction between the same
type of nodes (Robinson and Henderson 2018).Associated genes that are
targets of the same miRNAs have been found with help of correlation of
gene expression changes in recent algoritm (Markus List 2017). List et
al.~have specified that their approach can be useful for ceRNA studies
and published their approach as an R package.

@@

\hypertarget{mirnatarget-ratio-theoretical-action-in-network-based-cerna-model}{%
\subsubsection{miRNA:target ratio theoretical action in network based
ceRNA
model}\label{mirnatarget-ratio-theoretical-action-in-network-based-cerna-model}}

We developed a novel network based approach that can use to simulation
of gene regulation after any perturbation at miRNA:target interactions.
If a perturbation or change induces on graph, whole interacted nodes
vary because of changes in miRNA:target ratios. The system was assumed
as steady state that miRNAs affect according to the ratio of the
individual target to all targets (\autoref{fig1}a, expression values
shown as integer for comprehensible sampling). When expression of gene 2
(G2) increased to two fold (\autoref{fig1}b), distribution of miRNA (M1)
expression changes (\autoref{fig1}c). Consequently, expression of common
target (G4) of miRNA1 and miRNA2 also varies. The change at expression
of common target triggers to changes of proportional distribution of
miRNA2 (\autoref{fig1}d).

\begin{figure}
\hypertarget{fig1}{%
\centering
\includegraphics{Fig1.jpg}
\caption{Schematic presentation of mechanism of network based model. a)
In steady state, miRNAs (M: gray triangle) repress targets (G: circle
coconut) according to proportion of their targets' expression. b-c-d)
MiRNA expression distributions to targets vary when expression of one of
the targets perturbed.}\label{fig1}
}
\end{figure}

The model was developed by making some assumptions. Transcription rates
and degradation rates of miRNAs were assumed as stationary and equal.
Targets also have stable transcription and degradation rates, at the
same time they are equal. The factors that could be affect the
miRNA:target interaction are disregarded in this approach. It is
considered that miRNAs degrades or inhibits them via any mechanisms
after binding to target while the miRNAs were used reversibly. In
initial conditions, gene expression amounts in system have been
considered after repression activity of miRNAs. If a change occurs in
the system, variation on miRNA proportional distribution are reflected
as change of repression activity. \begin{equation} 
    Eff_{gi}= C_m * C_{gi}/\sum_{1}^{i} C_g \tag{1}\label{eq:1}
\end{equation} \begin{equation}
   R_{gi}= Eff_{i1}+ Eff_{i2} ... \tag{2}\label{eq:2}
\end{equation}

The efficiency of a miRNA on the individual target (\(Eff_{gi}\)) is
calculated according to equation \eqref{eq:1}; \(C_m\), miRNA expression
in the group; \(C_{gi}\), symbolize individual gene expression; \(C_g\)
gene expressions in the group (Groups are determined depending on
miRNAs. Targets that are interacted with the same miRNA find in a
group.). In this approach, because a gene can be found in different
groups, the cooperative activity of microRNAs also can be calculated.
Total repression activity of different miRNAs on the common target,
\(R\), is calculated based on equation \eqref{eq:2}; sum of efficiency
of different miRNAs on the target.

\begin{figure}
\hypertarget{fig2}{%
\centering
\includegraphics{fig2.jpg}
\caption{Calculations to determine of miRNA binding and repression
efficiency. \(G\), Gene; \(M\), miRNA; \(STE\), seed type effect;
\(RE\), Region Effect; \(E\), Energy; \(STE’\), normalized values of
seed type efficiency coefficient; \(RE’\), normalized values of region
efficiency coefficient; \(E’\), normalized values of energy
coefficient.}\label{fig2}
}
\end{figure}

\hypertarget{mirna-action-in-network-based-multifactorial-cerna-model}{%
\subsubsection{miRNA action in network based multifactorial ceRNA
model}\label{mirna-action-in-network-based-multifactorial-cerna-model}}

Interactions between miRNAs and their targets can be affected from
different parameters. For this reason, we improved model with factors
that can affect to miRNA activity. In our approach, the parameters are
classified as binding and efficiency factors and all of these are
evaluated in their groups. Binding factors determine interaction between
miRNA and target. The efficiency factors decide functionality of
binding. In the literature, binding free energy (Cao and Chen 2012 ;
Helwak et al. 2013) and seed type (Werfel et al. 2017) in miRNA:target
interactions are represented as binding affinity. Efficiency factors
determine how many of microRNA:target complexes will result in
inhibition. For example, binding region on the target shows whether the
target can be inhibited (functionality)(Hausser et al. 2013; Helwak et
al. 2013). Both of binding and efficiency factors are evaluated in the
groups and normalized according to equation \eqref{eq:3}; each factor is
normalized based on maximum variables that found in their groups.
\begin{equation}
F`= F/F_{max} \tag{3}\label{eq:3}
\end{equation}

The normalized values of factors take into account to determine binding
activity and miRNA efficiency on targets (\autoref{fig2}). Binding
affinities (activity, \(Eff\)) of miRNAs on each individual gene are
calculated as shown in equation \eqref{eq:4}; \(C_m\), miRNA expression
in the group; \(C_g\), Gene expression; \(C_{gi}\), individual gene
expression; gi, individual gene; g, whole genes in a group
(\autoref{fig2}c). \begin{equation}
Eff_{gi}= C_m * E`_{gi}*STE`_{gi}*C_{gi}/(\sum_{1}^{i} E`_{gi}*STE`_{gi}*C_{gi}) \tag{4}\label{eq:4}
\end{equation} \begin{equation}
Eff_{gi}= Eff_{gi}*RE`_{gi} \tag{5}\label{eq:5}
\end{equation}

After miRNA binds to its target, but might not repress to bound target.
The functionality of bound miRNA on target depends on efficiency factors
like region that is binding sequence of miRNA on its target. Exact
repression efficiency of miRNA is calculated according to equation
\eqref{eq:5} (\autoref{fig2}d); \(RE’_{gi}\), normalized values of
region efficiency coefficient between miRNA and gene. The cooperative
repression activity of miRNAs to their common targets is figured out as
shown in \autoref{fig2}e.

\hypertarget{application-of-model-in-a-breast-cancer-patient-dataset}{%
\subsubsection{Application of model in a breast cancer patient
dataset}\label{application-of-model-in-a-breast-cancer-patient-dataset}}

To determine suitability of method, we have tried to apply model in a
real dataset. Firstly, we obtained the miRNA and gene expression
datasets of the patient's tumor and normal tissue. We downloaded the
high-throughput experimental datasets which are provided miRNA:gene
target pairs with interaction factors by
\textbf{moore\_mirnatarget\_2015} and (Helwak et al. 2013). We have
combined miRNA and gene expression datasets via miRNA:target gene
dataset \textbf{(Whole steps can be found in Supplementary
data(TCGA\_E9-A1N5\_article.Rmd)}). We selected a gene which is over
expressed in tumor tissue with compared to normal tissue. It is pointed
out that ABCC1 gene, \emph{Multidrug resistance-associated protein 1},
is one of the most significant factor to develop resistance against
chemotherapotic agents(Atalay, Demirkazik, and Gunduz 2008; Lu et al.
2015; Atalay et al. 2006). So, we selected ABCC1 gene for simulation of
integrated dataset. We have detected the iteration that is suitable for
ABCC1 gene simulation. After simulation of system via ABCC1 gene with
over expression, we have compared simulation results and tumor tissue
expression levels.

@@

We have developed a model approach and workflow for competitive ceRNA
regulation in this study. The basic mode of miRNA repression activity
has been based on miRNA and target abundance in various researches(Arvey
et al. 2010; Denzler et al. 2014). For this reason, we performed first
application as considering only abundance of miRNAs and targets
(\autoref{fig1}). After the increasing in a gene expression value,
expression values of other genes also changed slightly in comparison
with alteration in perturbation started gene node. Different studies
have shown that the other gene expression values also rise differently,
if a gene abundance increases in ceRNA system (Lai, Wolkenhauer, and
Vera 2016; Salmena et al. 2011; Tay, Rinn, and Pandolfi 2014). It was
observed that primary neighborhoods of the trigger gene change with the
same ratio, but cooperative efficiency of miRNAs on common target causes
differently change in common target (\autoref{fig1}d). The common target
in system has displayed like a trigger for the other group and induced
changes of expression values of genes on the other group. Therefore,
effect of primary change of the gene (G2) on genes of the other group
was observed slightly (\autoref{fig1}d). In addition, like shown by
ceRNA hypothesis model of Ala et al., after the increasing of a gene
expression (G2), the miRNA that is found in the same group (M1) tended
to be more repressive on this target. This has caused to decrease in
increased expression value of primer triggering gene. When the
expression of a gene (G2) was increased to two fold, we observed that
the expression changes were more evident. So, we have considered that
high miRNA:target ratio results in stronger miRNA repression activity.
This results also are coherent with previous reports that were offered
by Arvey et al.~and Denzler et al.(Arvey et al. 2010; Bosson, Zamudio,
and Sharp 2014; Denzler et al. 2014). When the simulation was run in
sample dataset that includes lower target abundance, the first response
of primary neighborhood of the same group was determined as similar with
in system that have high target abundance. But regulations of ceRNAs
were observed differences in the following steps of simulation. In
overall state in the end of simulation, we have established that all
regulations more prominent when compared with low miRNA:target ratio
dataset. The model has appeared the importance of mRNA:target ratio in
regulation of competing RNAs.

@@

It has been reported in several studies that miRNA regulatory
interactions also is affected different parameters. For example, Xu et.
al have investigated the importance of seed pairing type between miRNAs
and their targets and target site location by using proteomics dataset
(Xu, Wang, and Liu 2014). They have proposed that the features of
binding between miRNA and target can be critical efficiency of miRNA and
can be used to find effective miRNA molecules. In addition, binding
energy between miRNA and targets also could be determinant for miRNA
efficiency. It has been reported by Breda et. al.~that strength of
miRNA:target interactions is depended on binding energy of complexes
(Breda et al. 2015). As similar to these, Bosson et al.~have associated
affinity and seed pairing of miRNA target pairs and suggested that
number of canonical base pairing is correlated with affinity. For these
reasons, we have improved the model with optional interaction parameters
that could be useful for understanding of miRNA repression activity. We
have explained the model in the same sample dataset with using different
factors and simulated via the same change (\autoref{fig3}). When the
factors were taken into account in the system, miRNA efficiencies varied
as shown in \autoref{fig3}a. Although the miRNA target ratios in initial
states were same in comparison with the sample dataset without factors,
efficiency of binding and repression have changed. When expression of a
gene (G2) increased, expression values of all genes also change
differentially because of contribution of efficiency factors
(\autoref{fig3} b,c,d). We have considered that in our approach energy
and seed type of pairs is significant for binding and targeted region is
important for repression. But these factors can be handled optionally
and differently.

In regulation of miRNA target sample system (\autoref{fig3}), miRNA (M1)
repressive efficiency on the primary triggering gene (G2) is low in
steady-state. So it has been observed that the regulatory activities of
miRNA efficiencies on the targets are weak after the increase of miRNA
target G2. When we ran to model with two fold increasing expression
value of common target (G4), the changes of other gene expressions have
observed more prominent. Furthermore, folding of expression of target
gene that has strong miRNA repression efficiency also resulted in
evident perturbation. On the other hand, It was observed that Gene2 was
weakly affected from change of Gene4 expression because of its weak
interaction factors.This also shows that the perturbation-initiation
element in the system is important parameter. So, we also have developed
an approach that is found perturbation efficiencies of elements in
dataset. \textbf{(see details in Supplementary File(fig1\_2app.Rmd)}.
When we applied the method on the minimal dataset, while Gene4 is the
most efficient element in terms of number of perturbed elements, it have
observed that Mir2 is most potent element in the mean of expression
change. In our approach, the elements behave differently based on type
(competing or miRNA). When a competing element was changed, whole
elements out of miRNAs are re-regulated. But, if a miRNA was perturbed,
whole competing elements are re-regulated and miRNA preserves expression
at after perturbation.

>>>>

  \title{Network based multifactorial modelling microRNA-target interaction.}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Selcen Ari\footnote{Department of Bioengineering, Yildiz Technical
  University, Istanbul, Turkey} \\ Alper Yilmaz\footnote{Department of Bioengineering, Yildiz Technical
  University, Istanbul, Turkey}}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{09 Ağustos 2019}

@@

Competing endogenous RNA (ceRNA) regulations and crosstalk between
various types of non-coding RNA in human is remarkable in means of miRNA
regulation. Many studies have pointed out that an alteration in
miRNA:target interaction can result in unexpected changes due to
indirect and complex interactions. In this paper, we defined a new
network-based model that handles miRNA:ceRNA interactions with
expression values. Our model is able to handle miRNA interaction factors
such as seed type, binding energy, if provided. Our approach is able to
reveal that a perturbation in an element of network affects whole
competing elements differently and cooperative efficiencies of miRNAs on
common targets could be calculated. Our findings emphasized importance
of miRNA:target ratios being crucial, as reported by previous studies.
We have showed that the competing elements which have the same or close
expression values may not be affected equally from the perturbation
because of repression functionality depended on interaction factors of
miRNA target pairs (TOFIX: Hocam etkileşim faktörlerinden dolayı
ekspresyonları yakın olsa bile baskılanma aktitesinin aynı
sergilenmeyeceğini ifade etmeye çalıştım. Bu şekilde anlaşılıyor mu?).
We applied the model to real sample consisting of breast cancer gene and
miRNA expression dataset and experimental miRNA:target interaction
dataset all generated via high throughput sequencing methods. A gene
over-expressed in tumor tissue, namely \emph{ABCC1}, is used as
perturbing element. We have observed that change in expression level of
single gene in miRNA:target network is sufficient to perturb regulations
in whole network, due to unforeseen and unpredicted regulation which are
only visible when considered in network context. Therefore, this model
helps unveiling the crosstalk between elements in miRNA:target network
where abundance of target and sponge effect are taken into account. The
model is scalable and can be plugged in with emerging miRNA effectors
such as circRNAs. The model is available as R package
\href{https://github.com/selcenari/ceRNAnetsim}{ceRNAnetsim}.

@@

As it is possible for miRNAs to suppress multiple targets, an individual
mRNA molecule can also be targeted by multiple miRNAs. In that case, the
targeted mRNAs exhibit competitor behavior, that is hypothesized as
competing endogenous RNAs (ceRNAs) (Ala et al. 2013; Cesana and Daley
2013), against their miRNAs. Briefly, Ala et al.~have explained the
ceRNA hypothesis as disturbance of the other target when one of the
targets on a steady-state system that included one miRNA and two target
was perturbed with expression change (Ala et al. 2013). Regarding
interaction between miRNAs and their target in a cell, explaining and
predicting results of an individual perturbation is difficult due to
complexity of interactions. Various computational and experimental
studies have tackled the problem of unraveling ceRNA:miRNA interactions.
It has been observed that when abundance of one of the targets of
miR-122 was increased, the other target expression also slightly
increased as a result of decreasing repressive activity of miR-122 on
them (Denzler et al. 2014). Bosson et al.~have developed a mathematical
model for changes on total target pool concentration after grouping
targets according to affinity and demonstrated that miRNA activity
correlated with affinity between miRNA and target (Bosson, Zamudio, and
Sharp 2014). Cooperative efficiency of miRNAs as well as competitor
behaviors of targets were also studied and it has been demonstrated to
be crucial for regulating available mRNA levels of targets (Denzler et
al. 2016). MiRNA:target interactions have been modeled as stoichiometric
and catalytic mechanisms and Figliuzzi et al.~have recommended handling
models in network context (Figliuzzi, Marinari, and De Martino 2013).
The model that can explain miRNA target interaction through topological
features has been applied at bipartite network by Nitzan et al.~(Nitzan
et al. 2014). Robinson and Henderson applied the model that handles
miRNA:target direct and indirect interactions via common miRNA of gene
and target of miRNA, at bipartite network. It has been demonstrated that
all miRNAs and targets in the network can interact with each other
through common miRNAs and genes, without interaction between the same
type of nodes (Robinson and Henderson 2018).Associated genes that are
targets of the same miRNAs have been found with help of correlation of
gene expression changes in recent algorithm (Markus List 2017). List et
al.~have specified that their approach can be useful for ceRNA studies
and published their approach as an R package.

@@

\hypertarget{construction-of-mirnatarget-network}{%
\subsubsection{Construction of miRNA:target
network}\label{construction-of-mirnatarget-network}}

miRNA and target pairs per line should be provided as edge list to
construct the network. At each line minimum required information is
expression levels of miRNA and the target. If available, additional data
about factors effecting binding or efficiency of miRNA can be provided
as separate columns. After construction of the network, amount of miRNA
per target is calculated and kept as edge data. Simply, a target will
sequester miRNA proportional to its ratio amount among other targets. If
additional criteria effecting the binding of miRNA to its target is
provided, distribution of miRNA will be calculated accordingly. Target
can be mRNA or any other ceRNA (circRNA, ncRNA, etc.) thus, throughout
the manuscript terms target, gene and ceRNA are used interchangeably.

\hypertarget{triggering-perturbation-and-subsequent-calculations}{%
\subsubsection{Triggering perturbation and subsequent
calculations}\label{triggering-perturbation-and-subsequent-calculations}}

Initially, the network is assumed in steady-state (\autoref{fig1}a)
condition and needs a trigger for initiating calculations. The trigger
can be a change in expression level of one or more genes
(\autoref{fig1}b). After a trigger, the network undergoes iterative
cycle of calculations at each of which distribution of miRNA in local
neighborhood is recalculated (\autoref{fig1}c). Based on new miRNA
distribution, expression level of each node (i.e.~ceRNA) is updated. Due
to nodes which are targeted by multiple miRNAs, the change in one
neighborhood spreads to other neighborhoods (\autoref{fig1}d),
consequently have potential to effect whole network due to ``ripple
effect''.

\begin{figure}
\hypertarget{fig1}{%
\centering
\includegraphics{Fig1.jpg}
\caption{Schematic presentation of mechanism of network based model. a)
In steady state, miRNAs (M: gray triangle) repress targets (G: circle
coconut) according to proportion of their targets' expression. b) Two
fold increase in transcript level of Gene2 (G2) acts as a trigger. c)
Distribution of miRNA1 (M1) changes. d) The change at expression of
common target effects changes of proportional distribution of miRNA2
(M2). Expression values are rounded to integers for
simplicity}\label{fig1}
}
\end{figure}

During calculations, following assumptions were adopted; 1)
Transcription and degradation rates of miRNAs are steady and equal. 2)
All available miRNAs are recycled as in miRNA:ceRNA binding, target is
degraded and miRNA is unaffected. 3) ceRNA targets also have stable
transcription and degradation rates and these rates are equal.

The repression efficiency of a miRNA on the individual target
(\(Eff_{gi}\)) is calculated according to equation \eqref{eq:1}; where
miRNA expression (\(C_m\)) in local neighborhood is distributed among
targets using individual gene expression levels (\(C_{gi}\)) (\#TODO
isn't \(C_g\) gene expressions in the group is equal to total of
\(C_{gi}\) in group?). For the genes targeted by multiple miRNAs,
cooperative activity of miRNAs, \(R\), is calculated by summing
repression activity of each miRNA (Equation \eqref{eq:2}).

\begin{equation} 
    Eff_{gi}= C_m \times C_{gi}/\sum_{1}^{i} C_g \tag{1}\label{eq:1}
\end{equation} \begin{equation}
   R_{gi}= Eff_{i1} + Eff_{i2} ... \tag{2}\label{eq:2}
\end{equation}

\hypertarget{multifactorial-calculations-in-mirnatarget-network}{%
\subsubsection{Multifactorial calculations in miRNA:target
network}\label{multifactorial-calculations-in-mirnatarget-network}}

Interactions between miRNAs and their targets can be affected from
various factors. So, our model integrates multiple factors when
calculating overall miRNA activity. We classified factors into two
categories. Factors effecting binding determine interaction between
miRNA and target and they alter amount of miRNA sequestered to target.
Factors effecting efficiency determine proportion of target mRNA
degraded by bound miRNA (\#TODO fix this sentence). In other words,
binding factors exert their influence before or during binding,
efficiency factors exert their influence after binding. In the
literature, binding free energy (Cao and Chen 2012 ; Helwak et al. 2013)
and seed type (Werfel et al. 2017) in miRNA:target interactions are
described as factors effecting binding affinity. Efficiency factors
determine how many of miRNA:target complexes will result in inhibition
and binding region on target drastically effect miRNA degradation
efficiency (Hausser et al. 2013; Helwak et al. 2013). Both binding and
efficiency factors are normalized to their maximum values and scaled to
{[}0,1{]} interval. The normalized values of factors take into account
to determine binding activity and miRNA efficiency on targets
(\autoref{fig2}). Binding affinities (activity, \(Eff\)) of miRNAs on
each individual gene are calculated as shown in equation \eqref{eq:4};
\(C_m\), miRNA expression in the group; \(C_g\), Gene expression;
\(C_{gi}\), individual gene expression; gi, individual gene; g, whole
genes in a group (\autoref{fig2}c). \begin{equation}
Eff_{gi}= C_m \times E^\prime_{gi} \times STE^\prime_{gi} \times C_{gi}/(\sum_{1}^{i} E^\prime_{gi} \times STE^\prime_{gi} \times C_{gi}) \tag{4}\label{eq:4}
\end{equation} \begin{equation}
Eff_{gi}= Eff_{gi}*RE^\prime_{gi} \tag{5}\label{eq:5}
\end{equation}

\begin{figure}
\hypertarget{fig2}{%
\centering
\includegraphics{fig2.jpg}
\caption{Calculations to determine of miRNA binding and repression
efficiency. \(G\), Gene; \(M\), miRNA; \(STE\), seed type effect;
\(RE\), Region Effect; \(E\), Energy; \(STE^\prime\), normalized values
of seed type efficiency coefficient; \(RE^\prime\), normalized values of
region efficiency coefficient; \(E^\prime\), normalized values of energy
coefficient.}\label{fig2}
}
\end{figure}

\iffalse

\%\% (\#TODO the equation is not considering binding factors, right?. We
should explain the calculation made in simulate function here, in simple
sentences. the function looks like this: \%\%
input\_graph\%\textgreater{}\% \%\%
tidygraph::activate(edges)\%\textgreater{}\% \%\%
group\_by(to)\%\textgreater{}\% \%\% mutate(mirna\_count\_per\_comp =
mirna\_count\_current\emph{comp\_count\_current}afff\_factor/sum(comp\_count\_current\emph{afff\_factor),
mirna\_count\_per\_comp = ifelse(is.na(mirna\_count\_per\_comp), 0,
mirna\_count\_per\_comp))\%\textgreater{}\% \%\%
ungroup()\%\textgreater{}\% \%\% mutate(effect\_pre = effect\_current,
effect\_current =
mirna\_count\_per\_comp}degg\_factor)\%\textgreater{}\% \%\%
group\_by(from)\%\textgreater{}\% \%\% mutate(comp\_count\_pre =
ifelse(comp\_count\_current\textless{} 0, 1, comp\_count\_current),
comp\_count\_current = comp\_count\_pre -
(sum(effect\_current)-sum(effect\_pre)), comp\_count\_current =
ifelse(comp\_count\_current\textless{} 0, 1,
comp\_count\_current))\%\textgreater{}\% \%\%
ungroup()\%\textgreater{}\% \%\% mutate(comp\_count\_list =
pmap(list(comp\_count\_list, comp\_count\_current), c), effect\_list =
pmap(list(effect\_list, effect\_current), c), mirna\_count\_list =
pmap(list(mirna\_count\_list, mirna\_count\_current), c)) \%\% ) \fi

After miRNA binds to its target, but might not repress to bound target.
The functionality of bound miRNA on target depends on efficiency factors
like region that is binding sequence of miRNA on its target. Exact
repression efficiency of miRNA is calculated according to equation
\eqref{eq:5} (\autoref{fig2}d); \(RE^\prime_{gi}\), normalized values of
region efficiency coefficient between miRNA and gene. The cooperative
repression activity of miRNAs to their common targets is figured out as
shown in \autoref{fig2}e.

\hypertarget{breast-cancer-patient-dataset}{%
\subsubsection{Breast cancer patient
dataset}\label{breast-cancer-patient-dataset}}

We have applied our model in a real dataset for which experimental
measurements of various factors were available. Expression levels of
miRNA and genes in tumor and normal tissue of single patient are
retrieved from TCGA(\#TODO cite). High-throughput experimental datasets
which are provided miRNA:gene target pairs with interaction factors
(Helwak et al. 2013; Moore et al. 2015). We have combined miRNA and gene
expression datasets via miRNA:target gene dataset retrieved from
\ldots{} (\#TODO name of database and its
citation)-\textgreater{}suplementary datasets of (Helwak et al. 2013;
Moore et al. 2015) . Detailed description of network construction and
its code is available in Supplementary data (\#TODO burada link nasıl
olacak, word dosyası veya PDF dosyası ismi mi vermemiz mi gerekiyor,
yoksa sadece Sup Data denilmesi yeterli mi? ilgili dosya:
TCGA\_E9-A1N5\_article.Rmd). ABCC1 gene, over-expressed in tumor tissue,
was selected as trigger for simulation of integrated dataset. After
simulation of network, we have compared simulation results and tumor
tissue expression levels.

@@

\hypertarget{networks-with-single-factor}{%
\subsubsection{Networks with single
factor}\label{networks-with-single-factor}}

We have developed a network-based approach to assess effects of
expression level changes in competitive ceRNA regulation. The basic mode
of miRNA repression activity has been based on miRNA and target
abundance in various researches(Arvey et al. 2010; Denzler et al. 2014).
Our approach can effortlessly calculate effects of expression changes
when abundance levels of miRNAs and targets is only available factor. In
sample network given in Figure (\autoref{fig1}), after an increase in
expression level of a gene (G2), expression values of other genes also
changed due to redistribution of miRNA among its targets. Previous
studies have shown that if a gene abundance increases in ceRNA system,
expression levels of genes targeted by shared miRNA are also affected
(Lai, Wolkenhauer, and Vera 2016; Salmena et al. 2011; Tay, Rinn, and
Pandolfi 2014). It was observed that primary neighborhoods of the
trigger gene change with the same ratio, but cooperative efficiency of
miRNAs on common target causes differently change in common target
(\#TODO last sentence needs some clarification, after ``but
cooperative\ldots{}'') (\autoref{fig1}d). Genes targeted by multiple
miRNAs act as a trigger for adjacent local neighborhood of targeting
miRNAs, causing changes in expression levels of genes outside the local
neighborhood of original trigger gene. Therefore, primary expression
level change in gene (G2) causes changes in other group of genes (G5 and
G6) even though original trigger gene (G2) and genes in other group are
not targeted by common miRNA. In addition, as shown by ceRNA hypothesis
model of Ala et al., after the increase of gene expression level of G2,
the miRNA that is found in the same group (M1) tended to be less
repressive on its remaining targets (G1, G3 and G4). It's important to
note that the changes in gene expression levels will have more
pronounced effect if miRNA:target ratio is high, i.e., more miRNA
available per target, which was reported in previous findings (Arvey et
al. 2010; Bosson, Zamudio, and Sharp 2014; Denzler et al. 2014).
\href{}{}

@@

\hypertarget{competing-endogenous-rnatarget-networks-based-on-interaction-factors}{%
\subsubsection{Competing Endogenous RNA:Target networks based on
interaction
factors}\label{competing-endogenous-rnatarget-networks-based-on-interaction-factors}}

\href{}{\#TODO}: \textless{}\textgreater{} ``title should be more
informative''-\textgreater{}

Earlier studies reported that miRNA regulatory interactions are affected
by different parameters. For example, Xu et al.~have investigated the
importance of seed pairing type between miRNAs and their targets and
target site location by using proteomics dataset (Xu, Wang, and Liu
2014). They have proposed that the features of binding between miRNA and
target can be critical for miRNA efficiency. In addition, binding energy
between miRNA and targets is a significant determinant for miRNA
efficiency and it has been reported that strength of miRNA:target
interactions is depended on binding energy of complexes (Breda et al.
2015). Similarly, another study revealed that affinity is correlated
with seed pairing of miRNA:target pairs and suggested that affinity is
correlated with number of canonical seed base pairing (Bosson, Zamudio,
and Sharp 2014). Therefore, we have integrated aforementioned
interaction parameters that could be useful for predicting miRNA
repression activity more accurately. The sample dataset used in Figure
\autoref{fig1} is recalculated with additional factors in effect
\href{}{} with same trigger, two fold increase in Gene2
(\autoref{fig3}). When the factors were taken into account in the
system, miRNA efficiencies varied as shown in \autoref{fig3}a. Although
the miRNA:target expression ratios in steady-state were same in
comparison with the sample dataset without factors, efficiency of
binding and repression have changed (compare miRNA distribution values
between Figure \autoref{fig1}a and Figure \autoref{fig3}a). When
expression of Gene2 (G2) increased, expression values of all genes also
changed differentially because of contribution of efficiency factors
(\autoref{fig3} b,c,d). We have considered that in our approach energy
and seed type of pairs is significant for binding and targeted region is
important for repression. \href{}{}

knitr::kable(minsamp)

In regulation of miRNA target sample system (\autoref{fig3}), miRNA (M1)
repressive efficiency on the primary triggering gene (G2) is low in
steady-state. So it has been observed that the regulatory activities of
miRNA efficiencies on the targets are weak after the increase of miRNA
target G2. When model is triggered with two fold increase in expression
level of common target (G4), the changes of other gene expressions have
observed more prominent \href{}{}. Furthermore, change in expression
level of target gene that has strong miRNA repression efficiency
resulted in evident perturbation in network. On the other hand, it was
observed that Gene2 was weakly affected from change of Gene4 expression
because of its weak interaction factors. Since perturbation efficiency
of each gene is different, we have developed a function which screens
each gene in the network for their perturbation efficiencies.
\textbf{(see details in Supplementary File(fig1\_2app.Rmd)}. When we
applied the method on the minimal dataset, Gene4 has been found to be
the most efficient element in terms of number of perturbed elements and
miRNA2 (M2) has been found to be causing the highest mean in expression
changes. \href{}{}

\href{}{} We used our model to simulate a real dataset which contains
thousands of genes and hundreds of miRNAs. Our model can successfully
simulate perturbations in such large network despite complex behaviors
and struggle to reach steady-state \textbf{(see details in Supplementary
File(TCGA\_E9-A1N5\_article.Rmd)}. Simulations show that change in
expression level of single gene has potential to effect whole network,
perturbing almost all nodes. These observations are in accordance with
competing endogenous RNA hypothesis where genes targeted with many
common miRNAs subsequently transmit perturbation to neighboring groups.

Network based approaches for analyzing miRNA:target interactions have
been developed in earlier studies. An initial attempt demonstrated the
ceRNA crosstalk in a network-like minimal interaction structure with
concentrations of ceRNA and miRNAs (Figliuzzi, Marinari, and De Martino
2013). Next, a network based kinetic model integrating miRNA and target
rates of transcription, degradation, binding and unbinding was developed
(Nitzan et al. 2014) using high throughput experimental dataset about
miR-92a depletion (Helwak et al. 2013). In this study, it was
demonstrated that distant ceRNAs can interact with each other via
indirect links, and the interactions are effected depending on distance
between ceRNAs or topological features of network (Nitzan et al. 2014).
More recently, an approach to detect ceRNA interaction by using the
miRNA expression, gene expression and common miRNAs between gene targets
was developed (Markus List 2017) which was effective in analyzing genes
through miRNAs. Based on the observations that a miRNA can exhibit
strong functionality to a target but may not against an other, the
authors have concluded that common miRNA based approach may not be
suitable for understanding regulations of ceRNA interactions.
